---
title: "Class 11: Genome Informatics"
author: "Kira"
format: pdf
---

## Section 1. Proportion of G/G in a Population

We can now read a CSV file that we downloaded from Ensemble.

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
# How many of each genotype based on total number of individuals (rows)
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

We might want to compare the proportion of SNPs to another population (GBR).

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378 (1)(gbr).csv")
head(gbr)
```

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100,2)
```

The proportion of individuals that are G|G is higher in the GBR population than in the MXL population.

## Section 4: Homework Questions

>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
# The total number of samples
nrow(expr)
```

```{r}
# To find the sample size for each genotype
table(expr$geno)
```

There are 108 individuals with genotype A/A, 233 individuals with genotype A/G, and 121 individuals with genotype G/G.

```{r}
# Putting the data into a data frame and pulling out relevant genotype G/G
expr.df <- data.frame(expr)
gg <- expr.df[expr.df$geno == "G/G",]
```

```{r}
# To find the median expression value for the G/G genotype
round(mean(gg$exp),2)
```

```{r}
# Repeating to select values for for A/A and A/G
ag <- expr.df[expr.df$geno == "A/G",]
aa <- expr.df[expr.df$geno == "A/A",]
```

```{r}
# Repeating to find the median expression value for A/A and A/G
round(mean(ag$exp),2)
round(mean(aa$exp),2)
```

The median expression for each genotype is as follows:

G/G : 20.59

A/G : 25.4

A/A : 31.82

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
ggplot(expr) + aes(x=geno,y=exp,fill=geno) + geom_boxplot(notch=TRUE) + xlab("Genotype") + ylab("Expression")
```

The expression of ORMDL3 decreases in the G/G genotype compared with the A/A genotype. We could hypothesize that the SNP influences expression but would need more detail/data to confirm this. 
